Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-02-17 paper.tex
Finish Section 6 besides "Limits," Improve testing, Update CUDA plot, Add citations, Fix graphics, Prepare to rework for OBE.

\begin{abstract}
  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained-Scans} to run on hardware without \emph{Forward-Progress Guarantees} (FPG) while avoiding deadlock. Additionally, we introduce a tile state representation for \emph{Chained-Scans} that does not rely on 64-bit atomics or memory fences for correctness, along with a subgroup-size-agnostic intra-workgroup implementation. On an FPG-lacking device---Apple M1 Pro (8+2)---\emph{Decoupled-Fallback} achieves near-\emph{memcopy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach. We further demonstrate the resilience of \emph{Decoupled-Fallback} against \emph{unfair} schedulers by emulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
\end{abstract}

@@

\subsection{Evolution of Inter-Workgroup Scan Architectures}
Contemporary GPU scans depart from early PRAM-like scans~\cite{Blelloch:1989:SAP} by leveraging the GPU memory hierarchy, which consists of progressively faster but increasingly private tiers. Efficient scans integrate fine-grained intra-workgroup (\emph{local}) strategies in shared memory and registers with coarse-grained inter-workgroup (\emph{global}) strategies in global memory. Given their low arithmetic intensity, local scans are inherently memory-bound, as demonstrated by Merrill and Grimshaw~\cite{Merrill2009}. Consequently, recent scan optimizations have shifted toward refining global strategies, particularly inter-workgroup coordination and synchronization. This section reviews three key approaches---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}---analyzing their handling of inter-workgroup dependencies and efficiency in minimizing global memory access while maximizing bandwidth utilization.

@@

\subsection{Ideal Scan Architecture}
\emph{Chained Scan} and \emph{Reduce then Scan} represent two contrasting paradigms in scan design: one embraces oversubscription and dynamic scheduling, while the other relies on a fixed-launch model with explicit control over occupancy. By avoiding explicit inter-workgroup synchronization, \emph{Chained-Scan} naturally aligns with the oversubscription paradigm, allowing the workgroup scheduler to dynamically manage occupancy. In contrast, \emph{Reduce-Then-Scan} employs workgroup raking, which mandates explicit occupancy control and, in some cases, inter-workgroup synchronization.

@@

Fixed-launch approaches, such as those used in \emph{Reduce-Then-Scan}, introduce several challenges. They require either careful tuning of workgroup launch sizes across different architectures or an occupancy discovery mechanism, which is nontrivial to implement. For example, the workgroup occupancy discovery technique proposed by Sorensen et al.~\cite{sorensen2016} relies on memory fences—unsupported in WebGPU—making it infeasible in certain environments.\footnote{Our implementation includes a fenceless occupancy discovery method suitable for WebGPU, but its description is beyond the scope of this work.}

@@

Hardware-level considerations further underscore the limitations of fixed-launches. On GPUs where the global memory data cache is indexed by physical addresses---and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)---fixed-launches can lead to performance degradation due to cache thrashing. This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups. Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns---or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations---may experience significant performance degradation.

Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture. By leveraging the workgroup scheduler, \emph{Chained-Scan} dynamically distributes work across architectures without requiring developer intervention. Furthermore, its single-kernel design and serial ordering make it suitable for \emph{in-situ} operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging.

However, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in both theoretical and practical terms, it has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that Apple’s M-series and ARM GPUs do not. Without forward progress guarantees, \emph{Chained-Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

\subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
A \emph{forward-progress guarantee} (FPG) ensures that all occupant subgroups (from possibly different workgroups) eventually make progress toward termination, preventing indefinite starvation. Sorensen et al.~\cite{sorensen2018,sorensen2021} provide the most comprehensive analysis of GPU schedulers, classifying \emph{Chained-Scan} as FPG-reliant. However, their work primarily focuses on formalizing progress models rather than their direct application to specific algorithms.

\raph{I think the deep-dive into the nature of deadlock can be trimmed back here, but chatting with John, such editing for concision can be done later, when we have a better sense about page count.}

The challenges faced by \emph{Chained-Scan} resemble the \emph{Dining Philosophers Problem}, where competing tasks share limited resources, risking deadlock. Coffman et al.~\cite{10.1145/356586.356588} define four conditions necessary for deadlock:
\begin{enumerate}
  \item \textbf{Mutual Exclusion}: Tasks claim exclusive control of resources they require.
  \item \textbf{Resource Holding}: Tasks hold acquired resources while waiting.
  \item \textbf{No Preemption}: A resource cannot be removed from a task before completion.
  \item \textbf{Circular Dependency}:
        \begin{enumerate}
          \item Dependency Relationships: A dependency exists such that each task holds one or more resources that are required by proceeding tasks.
          \item Circular Ordering: A circular chain of tasks exists.
        \end{enumerate}
\end{enumerate}
Applying this schema to \emph{Chained-Scan}, workgroups correspond to "tasks" and work-tile reductions to "resources." Conditions 1–3 hold due to the one-to-one mapping of work tiles to workgroups and the constraints of the GPU programming model. Condition 4a is satisfied by the inter-tile dependencies, leaving condition 4b---\emph{circular ordering}---as the key concern.

Although the serial assignment of work tiles ensures that predecessor tiles are scheduled or completed, it does not guarantee their reductions will be available when a workgroup begins \emph{Chained Propagation}. This uncertainty arises because workgroups are scheduled \emph{individually}, meaning a predecessor may not have had the opportunity to compute and post its reduction. In such cases, the dependent workgroup spins, consuming execution resources while waiting. Because workgroups may share execution resources by being co-resident on the same multiprocessor\john{Question, and maybe this is worth a footnote. What if we guaranteed that no multiprocessor ever had more than one workgroup scheduled to it, say by always allocating the maximum amount of workgroup memory? Of course there would be performance issues. But does that solve the problem? Or do we descend into an analogous issue with subgroup scheduling?}, without FPG, the scheduler may indefinitely prioritize the spinning workgroup over its predecessor, preventing progress and creating a circular dependency, satisfying condition 4b and resulting in deadlock.

\raph{Answering John's question here. Short answer no. One mental model that would illustrate this is a workgroup being preempted from a multiprocessor for system reasons, for example to perform compositing for scanout. I brought this up with a GPU engineer and they said this wasn't actually the reason for lack of FPG on their hardware, but wasn't able to explain the true story for proprietary reasons. In any case, while I am fine to a large extent talking operationally about scheduling etc, at the end of the day you need a formal model that tells you which computations will terminate. That's what Tyler's work is about and I think it's best to cite it and not try to re-litigate it, especially in oversimplified form. Also, taking another look at "Specifying and Testing," I notice that they use "livelock" but not deadlock to describe situations where system progress depends on FPG. I'm not finding a cite for this now, but if memory serves, a deadlock condition is one in which no progress is possible under *any* scheduling condition, while in livelock there is some scheduler condition that could resolve the livelock, but it is not resolved under all such conditions.} \john{The reason that I ask this question is that I take the perspective of a reviewer who decides, oh, I can solve this FPG problem by doing something simple, like ``scheduling only one workgroup per processor''. The reviewer, of course, doesn't actually implement this, but it totally kills the paper when a reviewer gets it into his/her head that there is another way to go about this and why didn't we study THAT way, reject, go back and incorporate my amazing idea. We need to explicitly close the door emphatically in the paper to obvious, simple, unworkable solutions; we can't give reviewers any headroom to substitute their own judgment when we know better.}

For example, consider a scan operation requiring two workgroups on a GPU with a single multiprocessor. Both workgroups are co-resident, but the multiprocessor has only one scheduling unit, which lacks FPG\@. Both workgroups launch and acquire their respective work tiles. The workgroup responsible for \emph{Work Tile 1} finishes first and begins spinning, repeatedly querying the memory address for the reduction of \emph{Work Tile 0}. However, because the scheduler lacks FPG, it continues scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction. As a result, \emph{Work Tile 1} spins indefinitely, potentially triggering \emph{timeout detection and recovery}\footnote{Operating systems have different mechanisms for handling unresponsive programs. We borrow here Window's terminology. On Windows, the Timeout Detection and Recovery system resets the GPU after 2 seconds by default. iOS has a timeout watchdog, which, interestingly, does not terminate a spinning kernel. On Linux, the behavior depends on the GPU driver.} (TDR) and crashing the host program.

@@

  $\textit{part\_id} \gets \textit{wg\_broadcast}$\;

  \textit{local\_vectors}: $\textnormal{array}\langle \textnormal{vec4}\langle u32\rangle, \textit{p} \rangle$\;

  \textbf{Load}($\textit{in}$, $\textit{part\_id}$, \textit{local\_vectors})\;
  \textbf{SubgroupRake}(\textit{local\_vectors})\;

  \If{$\textnormal{highestRankLane}(\textit{threadid})$}{
    $\textit{wg\_partials}[\textit{subgroup\_id}] \gets \textit{local\_vectors}[\textit{p} - 1]$\;
  }
  \textbf{barrier()}\;

@@

  \tcp{Chained Propagation, Decoupled Fallback}
  \While{$\textit{wg\_control} = \textnormal{LOCKED}$}{
    \textbf{Lookback}($\textit{spine}$, $\textit{wg\_prev\_reduction}$, $\textit{wg\_control}$)\;
    \textbf{barrier()}\;

@@

  \textbf{Write}($\textit{out}$, $\textit{part\_id}$, \textit{local\_vectors}, $\textit{wg\_partials}$, $\textit{wg\_prev\_reduction}$)\;

@@

\section{Implementation}
At a high level, our algorithm follows a five-phase construction:
\begin{enumerate}
  \item[(0)] \textbf {Initialization}: The first thread in the workgroup atomically acquires its work-tile by atomically incrementing a value in global memory.
  \item \textbf{Load Phase}: Each thread loads $p$ vectors, consuming the input in a subgroup-blocked ordering. For a subgroup size $s$ and vector size $v$, each subgroup processes a contiguous block of size $s \cdot p \cdot v$.
  \item \textbf{Subgroup-Rake}: Each thread scans its vectors privately, then joins $p$ raking Kogge-Stone~\cite{5009159} subgroup scans,
        allowing the subgroup-block scan to remain barrier-free and in registers. The subgroup-block reduction is then posted to shared memory. For a workgroup of size $w$, subgroup-raking reduces the size of the local spine to $w/s$.
  \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup partial reductions, writing the results back to shared memory upon completion.
  \item \textbf{Chained Propagation, Decoupled Fallback}: The first subgroup in the workgroup performs a lookback operation to acquire the reduction result of all previous workgroups, engaging in fallback operations as necessary. Once completed, this subgroup posts the results to shared memory for the workgroup to consume.
  \item \textbf{Write Phase}: Each subgroup incorporates the results from the workgroup-wide scan and lookback operation to produce and write out the correct output.
\end{enumerate}
Phase 0 ensures that the work tiles are consumed in serial order. Phases 1, 2, 3, and 5 comprise the intra-workgroup scan, following a \emph{Scan-then-Propagate} strategy. Phase 4, \emph{Decoupled-Fallback}, is responsible for managing all inter-workgroup dependencies. As our loading and writing phase are typical of scan kernels, we focus here on our portability-specific adaptations.

@@

\subsection{Workgroup-Wide Scan}
The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches. To handle this variability, we use a generalized radix-$s$ Ladner-Fischer~\cite{10.1145/322217.322232} scan network embedded with Kogge-Stone~\cite{5009159} subgroup scans to perform the scan across the local spine. Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts. To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$. Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction. Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index. However, such access is interpreted by the hardware as a subgroup \emph{broadcast} operation, resulting in no conflicts. Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase causes no performance degradation due to our modest overall use of shared memory, such as our deliberate retention of vectors in registers and the efficiency of subgroup raking. Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes.\footnote{Since we also require a workgroup-wide pure reduce operation, we obtain this by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}

@@

\subsection{Tile State Implementation}
In all variants of \emph{Chained-Scan}, there is communication and synchronization between workgroups. The fundamental primitive is \emph{message passing,} meaning that one workgroup posts the result of a communication, then other workgroups read it. This primitive is well studied in the context of memory models~\cite{}. \raph{The earliest publication of the MP litmus test appears to be section 2.1 of "Intel 64 Architecture Memory Ordering White Paper". I'm not sure that's the best cite; the MC mutants paper cites https://arxiv.org/pdf/1207.7264} The key requirement is both control synchronization, meaning the reader does not proceed until the writer sends the message; and data synchronization, meaning that the reader reads the correct value. The latter is especially challenging in relaxed memory models as are employed on GPUs.

Combined control and data synchronization can be accomplished by packing both the tile state and the reduced value into a single word, using relaxed atomic store to send the message, and relaxed atomic load to receive the message, spinning as necessary until the tile state is updated. The semantics of atomics guarantee, even with relaxed memory ordering, that the data is fresh. However, if the tile state and data are stored in two different memory locations, then relaxed atomics are not sufficient and some other sychronization is necessary. On CPUs, it is common to upgrade the memory semantics from relaxed to acquire/release, which then provides the necessary guarantee. \raph{Here's a good place to cite the MP litmus test because that's exactly what it's about.} Unfortunately, on GPUs, acquire/release semantics are neither portable (they are only available on Vulkan, and optional until version 1.3), nor are they necessarily efficient even when they are available; often the implementation involves expensive cache flushes. A device-scoped memory barrier has similar problems. While it would provide the necessary guarantee, it is also not available in WebGPU (or in versions of Metal prior to 3.2), and is expensive.

@@

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. Through testing, \john{across all devices (?)} \thomas{Only rigorously tested on the 2080, everything else was very light.}we find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@. For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum. Our test hardware, listed POSITIONAL-RELATION, spans a wide range of FPG and performance capabilities. We empirically find that the Google Chrome native implementation of WebGPU (Dawn) provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires Android native, necessitating the use of WGPU\@.\footnote{To keep the perforamance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
\begin{table}
  \centering
  \small
  \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} l l c r r}
    \toprule
    Vendor & Device Name & OBE & Occupancy & Subgroup Size \\
    \midrule
    Intel  & HD Graphics 620        & \textbf{Yes} & 9   & 8--32  \\
    ARM    & Mali-G78 MP20          & \textbf{Yes} & 40  & 16  \\
    Apple  & M3                     & \textbf{No}  & 80  & 32  \\
    Apple  & M1 Max                 & \textbf{No}  & 160 & 32  \\
    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192 & 32  \\
    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420 & 32--64  \\
    \bottomrule
  \end{tabular*}
  \caption{Hardware Specifications of Testing Devices.}
  \label{tab:hardware}
\end{table}

\subsection{Performance Comparison}
\subsubsection{Performance Versus Contemporary Implementations:}
We first compare \emph{Decoupled Fallback} (\emph{CSDLDF}) to the \emph{Decoupled Lookback} (\emph{CSDL}) implementation found in NVIDIA’s CUB library, using the 2080 Super only. To assess the impact of differing APIs, we also implement \emph{CSDLDF} in CUDA.\footnote{Due to WebGPU’s 128MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{graphics/cuda_plot.pdf}
  \caption{Performance versus CUB (NVIDIA 2080 Super)}
\end{figure}

@@

As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at a 100\% fallback rate across all devices—except for the Intel HD620, where only a 25\% fallback rate is induced. Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%. While Figures~\ref{fig:fallbacks_initiated} and \ref{fig:total_spins} confirm that we are inducing fallbacks, this test primarily measures the impact of increasing fallback operations rather than true blocking behavior. The finer details of this distinction are discussed next.
\section{Discussion}
\subsection{Limits on Speed-of-Light Performance}

>>>>

\begin{abstract}
  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained-Scans} to run on hardware without \emph{Forward-Progress Guarantees} (FPG) while avoiding deadlock. Additionally, we introduce a tile state representation for \emph{Chained-Scans} that does not rely on 64-bit atomics or memory fences for correctness, along with a subgroup-size-agnostic intra-workgroup implementation. On an FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled-Fallback} achieves near-\emph{memcopy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach. We further demonstrate the resilience of \emph{Decoupled-Fallback} against \emph{unfair} schedulers by emulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
\end{abstract}

@@

\subsection{Evolution of Inter-Workgroup Scan Architectures}
Contemporary GPU scans depart from early PRAM-like scans~\cite{Horn, Hensley, Sengupta2006, Gress} by leveraging the GPU memory hierarchy, which consists of progressively faster but increasingly private tiers. Efficient scans integrate fine-grained intra-workgroup (\emph{local}) strategies in shared memory and registers with coarse-grained inter-workgroup (\emph{global}) strategies in global memory. Given their low arithmetic intensity, local scans are inherently memory-bound, as demonstrated by Merrill and Grimshaw~\cite{Merrill2009}. Consequently, recent scan optimizations have shifted toward refining global strategies, particularly inter-workgroup coordination and synchronization. This section reviews three key approaches---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}---analyzing their handling of inter-workgroup dependencies and efficiency in minimizing global memory access while maximizing bandwidth utilization.

@@

\subsection{Ideal Scan Architecture}
\emph{Chained Scan} and \emph{Reduce-Then-Scan} represent two contrasting paradigms in scan design. By avoiding explicit barrier-based inter-workgroup synchronization, \emph{Chained Scan} aligns with the oversubscription paradigm, delegating occupancy management to the workgroup scheduler. In contrast, \emph{Reduce-Then-Scan} employs workgroup raking, requiring fixed launches and explicit occupancy control, forcing developers to manually tune launch sizes or implement occupancy discovery techniques.

Hardware-level considerations underscore the limitations of fixed-launches. On GPUs where the global memory data cache is indexed by physical addresses---and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)---fixed-launches can lead to performance degradation due to cache thrashing. This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups. Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns---or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations---may experience significant performance degradation.

Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture. Furthermore, its single-kernel design and serial ordering make it suitable for \emph{in-situ} operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging. However, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in both theoretical and practical terms, it has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that Apple’s M-series GPUs do not \thomas{I've cut mentioning ARM here due to reclassification, but it kind of reduces the impact. I.e. "Oh it's just Apple that's missing, so this is a vendor hack?"}. Without forward progress guarantees, \emph{Chained-Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

@@

@@

\thomas{Edited this section in response to above comments. Distinguished Chained Scan as specifically eliding barrier based sychronization. Linked fixed launch more clearly to manual tuning and occupancy discovery. Shortened section. OK to remove comments?}

\subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
A \emph{forward-progress guarantee} (FPG) ensures that all occupant subgroups (from possibly different workgroups) eventually make progress toward termination, preventing indefinite starvation. Sorensen et al.~\cite{sorensen2018,sorensen2021} provide the most comprehensive analysis of GPU schedulers, classifying \emph{Chained-Scan} as FPG-reliant.
\begin{itemize}
  \item Introduce Sorensen
  \item Introduce Progress models
  \item Explain why OBE + data
  \item Explain why one-to-one mapping doesn't solve problem.
\end{itemize}

\john{Question, and maybe this is worth a footnote. What if we guaranteed that no multiprocessor ever had more than one workgroup scheduled to it, say by always allocating the maximum amount of workgroup memory? Of course there would be performance issues. But does that solve the problem? Or do we descend into an analogous issue with subgroup scheduling?}

\raph{Answering John's question here. Short answer no. One mental model that would illustrate this is a workgroup being preempted from a multiprocessor for system reasons, for example to perform compositing for scanout. I brought this up with a GPU engineer and they said this wasn't actually the reason for lack of FPG on their hardware, but wasn't able to explain the true story for proprietary reasons. In any case, while I am fine to a large extent talking operationally about scheduling etc, at the end of the day you need a formal model that tells you which computations will terminate. That's what Tyler's work is about and I think it's best to cite it and not try to re-litigate it, especially in oversimplified form. Also, taking another look at "Specifying and Testing," I notice that they use "livelock" but not deadlock to describe situations where system progress depends on FPG. I'm not finding a cite for this now, but if memory serves, a deadlock condition is one in which no progress is possible under *any* scheduling condition, while in livelock there is some scheduler condition that could resolve the livelock, but it is not resolved under all such conditions.}

\john{The reason that I ask this question is that I take the perspective of a reviewer who decides, oh, I can solve this FPG problem by doing something simple, like ``scheduling only one workgroup per processor''. The reviewer, of course, doesn't actually implement this, but it totally kills the paper when a reviewer gets it into his/her head that there is another way to go about this and why didn't we study THAT way, reject, go back and incorporate my amazing idea. We need to explicitly close the door emphatically in the paper to obvious, simple, unworkable solutions; we can't give reviewers any headroom to substitute their own judgment when we know better.}

@@

  $\textit{tile\_id} \gets \textit{wg\_broadcast}$\;

  $\textit{local\_vectors}$ \gets $\textnormal{array}\langle \textnormal{vec4}\langle u32\rangle, \textit{p} \rangle$\;

  \textbf{Load}($\textit{in}$, $\textit{tile\_id}$, \textit{local\_vectors})\;
  \textbf{SubgroupRake}($\textit{local\_vectors}$, $\textit{wg\_partials}$)\;
  \textbf{barrier()}\;

@@

  \tcp{Chained Propagation, Decoupled Fallback}
  \textbf{PostLocalReduction}($\textit{spine}$, $\textit{tile\_id}$, $\textit{wg\_partials}$)\;
  \While{$\textit{wg\_control} = \textnormal{LOCKED}$}{
    \textbf{Lookback}($\textit{spine}$, $\textit{tile\_id}$, $\textit{wg\_prev\_reduction}$, $\textit{wg\_control}$)\;
    \textbf{barrier()}\;

@@

  \textbf{Write}($\textit{out}$, $\textit{tile\_id}$, $\textit{local\_vectors}$, $\textit{wg\_partials}$, $\textit{wg\_prev\_reduction}$)\;

@@

\section{Implementation}
At a high level, our algorithm follows a five-phase construction:
\begin{enumerate}
  \item[(0)] \textbf {Initialization}: The first thread in the workgroup atomically acquires its work-tile by atomically incrementing a value in global memory.
  \item \textbf{Load Phase}: Each thread loads $p$ vectors, consuming the input in a subgroup-blocked ordering. For a subgroup size $s$ and vector size $v$, each subgroup processes a contiguous block of size $s \cdot p \cdot v$.
  \item \textbf{Subgroup-Rake}: Each thread scans its vectors privately, then joins $p$ raking Kogge-Stone~\cite{5009159} subgroup scans,
        allowing the subgroup-block scan to remain barrier-free and in registers. The subgroup-block reduction is then posted to shared memory. For a workgroup of size $w$, subgroup-raking reduces the size of the local spine to $w/s$.
  \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup partial reductions, writing the results back to shared memory upon completion.
  \item \textbf{Chained Propagation, Decoupled Fallback}: The local reduction is posted to global memory. Then, the first subgroup in the workgroup performs a lookback operation to acquire the reduction result of all previous workgroups, engaging in fallback operations as necessary. Once completed, this subgroup posts the results to shared memory for the workgroup to consume.
  \item \textbf{Write Phase}: Each subgroup incorporates the results from the workgroup-wide scan and lookback operation to produce and write out the correct output.
\end{enumerate}
Phase 0 ensures that the work tiles are consumed in serial order. Phases 1, 2, 3, and 5 comprise the intra-workgroup scan, following a \emph{Scan-then-Propagate} strategy. Phase 4, \emph{Decoupled-Fallback}, is responsible for managing all inter-workgroup dependencies. As our loading and writing phase are typical of scan kernels, we focus here on our portability-specific adaptations.

@@

\subsection{Workgroup-Wide Scan}
The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches. To handle this variability, we use a generalized radix-$s$ Ladner-Fischer~\cite{10.1145/322217.322232} scan network embedded with Kogge-Stone~\cite{5009159} subgroup scans to perform the scan across the local spine. Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts. To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$. Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction. Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index. However, such access is interpreted by the hardware as a subgroup \emph{broadcast} operation, resulting in no conflicts. Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase causes no performance degradation due to our modest overall use of shared memory, such as our deliberate retention of vectors in registers and the efficiency of subgroup raking. Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes.\footnote{We may also obtain a workgroup-wide reduce operation by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}

@@

\subsection{Tile State Implementation}
In all variants of \emph{Chained-Scan}, there is communication and synchronization between workgroups. The fundamental primitive is \emph{message passing,} meaning that one workgroup posts the result of a communication, then other workgroups read it. This primitive is well studied in the context of memory models~\cite{Alglave0, MCMutants}. \raph{The earliest publication of the MP litmus test appears to be section 2.1 of "Intel 64 Architecture Memory Ordering White Paper". I'm not sure that's the best cite; the MC mutants paper cites https://arxiv.org/pdf/1207.7264} \thomas{Added MC Mutants + the Algave paper. OK? Or should cite more? Alglave has tons of papers.} The key requirement is both control synchronization, meaning the reader does not proceed until the writer sends the message; and data synchronization, meaning that the reader reads the correct value. The latter is especially challenging in relaxed memory models as are employed on GPUs.

Combined control and data synchronization can be accomplished by packing both the tile state and the reduced value into a single word, using relaxed atomic store to send the message, and relaxed atomic load to receive the message, spinning as necessary until the tile state is updated. The semantics of atomics guarantee, even with relaxed memory ordering, that the data is fresh. However, if the tile state and data are stored in two different memory locations, then relaxed atomics are not sufficient and some other sychronization is necessary. On CPUs, it is common to upgrade the memory semantics from relaxed to acquire/release, which then provides the necessary guarantee~\cite[Section 10.2]{intel_sdm_2024} \thomas{I've cited the latest publication by Intel. The section is nearly identical to 2007.} . Unfortunately, on GPUs, acquire/release semantics are neither portable (they are only available on Vulkan, and optional until version 1.3), nor are they necessarily efficient even when they are available; often the implementation involves expensive cache flushes. A device-scoped memory barrier has similar problems. While it would provide the necessary guarantee, it is also not available in WebGPU (or in versions of Metal prior to 3.2), and is expensive.

@@

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. We find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@. For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum. Our test hardware, listed POSITIONAL-RELATION, spans a wide range of FPG and performance capabilities. We empirically find that the Google Chrome native implementation of WebGPU (Dawn) provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires Android native, necessitating the use of WGPU\@.\footnote{To keep the perforamance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
\begin{table}
  \centering
  \small
  \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} l l c r r}
    \toprule
    Vendor & Device Name & OBE & Occupancy & Subgroup Size \\
    \midrule
    Intel  & HD Graphics 620        & \textbf{Yes} & 9   & 8--32  \\
    ARM    & Mali-G78 MP20          & \textbf{Yes} & 40  & 16  \\
    Apple  & M3                     & \textbf{No}  & 80  & 32  \\
    Apple  & M1 Max                 & \textbf{No}  & 160 & 32  \\
    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192 & 32  \\
    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420 & 32--64  \\
    \bottomrule
  \end{tabular*}
  \caption{Hardware Specifications of Testing Devices.}
  \label{tab:hardware}
\end{table}

\subsection{Performance Comparison}
\subsubsection{Performance Versus Contemporary Implementations:}
We first compare \emph{Decoupled Fallback} (\emph{CSDLDF}) to the \emph{Decoupled Lookback} (\emph{CSDL}) implementation found in NVIDIA’s CUB library, using the 2080 Super only. To assess the impact of differing APIs, we also provide the results of an identical \emph{CSDLDF} in CUDA.\footnote{Due to WebGPU’s 128MB storage buffer limit, our maximum test size is limited to $2^{25}$ 32-bit elements.}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{graphics/cuda_plot.pdf}
  \caption{(This graph can be provided with or without Memcpy speeds. I am thinking without, as it's currently very busy.) Performance versus CUB (NVIDIA 2080 Super)}
\end{figure}

@@

As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at near 100\% fallback rates across all devices---except for the Intel HD620, where only a 25\% fallback rate is induced. Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%. While Figures~\ref{fig:fallbacks_initiated} and \ref{fig:total_spins} confirm that we are inducing fallbacks, this test primarily measures the impact of increasing fallback operations rather than true blocking behavior. The finer details of this distinction are discussed next.
\section{Discussion}
\subsection{Limits on Speed-of-Light Performance}
